---
title: "Basic, Intermediate, and Advanced SQL in R"
author: "Abdulaziz AlYaqout"
date: "10/19/2018"
output:
  pdf_document: default
  html_document: default
---

## Introduction to SQL

Structured Query Language (or SQL) is a language that is used to communicate with databases. It is the primary language used for relational database systems. 

###SQL in R
This document walks through basic, intermediate, and advanced uses of SQL. 

For ease of use, I demonstrate all the SQL code in R. R contains packages that allow you to write queries in SQL. In some cases (like a work setting) you can connect directly to a database in R (e.g. through an ODBC connection). But for the sake of simplicity, in this document, rather than connecting directly to a database, we will simply import data into R data frames and show how to run SQL queries over it. 

The R dataframes will act as our database tables. 

We will use the sqldf package in R to do this.

```{r}
library(sqldf)
```


##Data

We will use data related to the real estate market in Kansas City, MO. 
```{r}
library(readxl)
kc_housing <- read_excel("KC_House_Data.xlsx")
```

####Explore structure of the data
```{r}
str(kc_housing)
```
Looks like most of our variables are of numeric type.

We can also aggregate by some factors (e.g. bedoroom size, waterfront, etc.)

##Basic SQL

####Simple Selects
To see all available fields in a table, run a 'SELECT *' command.
In the FROM clause, specify the dataframe name as the 'table' name. 
And we will limit our results to 10 records to save space.

```{r}
sqldf("SELECT * 
      FROM kc_housing 
      LIMIT 10")
```

If you want to select just a few fields, specify those in the SELECT clause instead of the *. 
```{r}
sqldf("SELECT id, price, bedrooms 
      FROM kc_housing 
      LIMIT 10")
```

####Select count
Another common SQL command is 'select count()' which returns a count of records. 

The following query returns a count of all records in the table. 
```{r}
sqldf("SELECT COUNT(*) 
      FROM kc_housing")
```


####Select distinct
Sometimes you may just want the unique values in a field. The following query
will show how many unique floor counts there are in the KC houses. 
```{r}
sqldf("SELECT DISTINCT floors 
      FROM kc_housing")
```

Integrating our knowledge of count() function, we can find the number of distinct
floors in the KC housing data. 

```{r}
sqldf("SELECT COUNT(DISTINCT floors) 
      FROM kc_housing")
```

Note: this is useful, but the field name is a little unclean. Let's clean it up.
You can rename a field directly using SQL code using 'as' for a cleaner output. We'll name it
'NumberOfDistinctFloors'

```{r}
sqldf("SELECT COUNT(DISTINCT floors) as NumberOfDistinctFloors
      FROM kc_housing")
```
#### WHERE Clause
In the WHERE clause, we filter the data. The SELECT tells us what fields to
return. The WHERE clause puts a filter on the data. 

Lets return the count (number) of all houses with more than 2 bathrooms
```{r}
sqldf("SELECT COUNT(*) as HighBathroomHouses
      FROM kc_housing
      WHERE bathrooms > 2")
```
And compare it to the 1-bathroom houses
```{r}
sqldf("SELECT COUNT(*) as OneBathroomHouses
      FROM kc_housing
      WHERE bathrooms =  1")
```

As you can see, the WHERE clause takes logical operators (<, >, =, and !=)

You can link multiple conditions together using AND and OR as well. 

Let's return 15 records of  2 bedroom houses whos living rooms are greater than 
2000 square feet. 
```{r}
sqldf("SELECT id, bathrooms, sqft_living
      FROM kc_housing
      WHERE bathrooms = 2 AND sqft_living > 2000
      LIMIT 15")
```
Now let's return 15 records of 2 or 3 bedroom houses whos living rooms are greater than 
2000 square feet and whose condition is at least a rating of 5. 
```{r}
sqldf("SELECT id, bathrooms, sqft_living, condition
      FROM kc_housing
      WHERE (bathrooms = 2 OR bathrooms = 3) AND sqft_living > 2000 AND condition >= 5
      LIMIT 15")
```

##Intermediate SQL
The two intermediate functions in SQL are to group by and to join tables. 

####Group by
Group by allows you to run the query aggregating the results on some factor.

Here we'll return the count of records in the table, grouping by condition

```{r}
sqldf("SELECT condition, COUNT(*) as Volume
      FROM kc_housing
      GROUP BY condition")
```

Suppose we want to order the results by Volume. We can order our results
by typing 'ORDER BY' and the variable to order by. 

```{r}
sqldf("SELECT condition, COUNT(*) as Volume
      FROM kc_housing
      GROUP BY condition
      ORDER BY Volume")

```

Group by is also useful when performing functions on data and comparing groups. 

To perform functions in SQL, you type the function name and call it on the field 
of interest. For example, to find the minimum yr built, type min(yr_built). 

```{r}
sqldf("SELECT MIN(yr_built)
      FROM kc_housing")

```
Now let's find the most recent year a house was built, grouping by condition
```{r}
sqldf("SELECT condition, MAX(yr_built) as MostRecentYear
      FROM kc_housing
      GROUP BY condition
      ORDER BY condition")

```

The worst quality houses (1 and 2) not surprisingly are older than the 
better quality houses. 

####Inner Joins
A key aspect of SQL is joining tables together. This is central to the relational
structure of databases. 

To illustrate, let's make some data that we can join to the housing data. 
We'll assume that each house owner owns a car. We'll take the cars from the
classic mtcars dataset and randomly assign them to each id in the KC housing data. 

```{r}
kc_cars <- data.frame(id = kc_housing$id,
                      cartype = sample(rownames(mtcars), nrow(kc_housing), 
                                       replace=TRUE))

head(kc_cars)
```
For each ID in KC Housing, there is an associated car. 

Suppose we want to find the cars in houses whose lots exceed 5000 square feet. 

We can join the KC Housing data with the KC Cars data via a common key
called a primary key. This binds the two tables together. In our case it is id.

Note: an inner join will only return records with ids that are common to both tables.

```{r}
sqldf("SELECT kc_housing.id, price, bedrooms, cartype
      FROM kc_housing
      INNER JOIN kc_cars
      ON kc_housing.id = kc_cars.id
      WHERE kc_housing.sqft_lot > 5000
      LIMIT 15")
```
Now you can join the two tables. The associated car type will be joined to the 
KC Housing on the records with a matching ID. This is essentiall the idea of joins.

####Left Joins and Right Joins
Suppose your cars data only contains a subset of the IDs in the housing data. 
But you still want to retain all the records in the housing data in your query.

A left join will allow you to do that. 

It retains all records of the left table, while performing a join with columns
on the right table where appropriate. 

Subset the kc_cars to just half the data and do a left join with kc_housing
```{r}
set.seed(42)
kc_cars_subset <- kc_cars[sample(1:nrow(kc_housing), nrow(kc_housing)*.5),]
```


```{r}
sqldf("SELECT kc_housing.id, price, bedrooms, cartype
      FROM kc_housing
      LEFT JOIN kc_cars_subset
      ON kc_housing.id = kc_cars_subset.id
      LIMIT 10")
```

As you can see many of the rows contain NA. Left joins retain all records
from the left table, and only matching records of the right table. Left join 
fills in NA or NULL where there is no matching key in the right table. 

Right join does the opposite to this (reversing left and right tables).

There is one more join called outer join. This returns all records in both tables.
It fills in NA on the left table where there is no match to the right table key. 
And NA on the right table where there is no match to the left table key. Sometimes
this is called a full outer join. 

##Advanced SQL

There are many advanced SQL techniques. Advanced SQL includes building nested queries, 
doing multiple joins, and building case statements to create variables in your table
using if/else logic. 

####Nested queries
The simple idea behind a nested query is to return records using SQL
statement, and then query from those results as if the results were its own table.

Here is a simple example. Suppose we want to find the maximum average 
square footage of a lot size grouping by condition. 

We could just do the following query which groups the data by condition
and finds the max square footage for each of these groups. 
We can then order by the MaxSqFootage in descending order. 

```{r}
sqldf("SELECT condition, MAX(sqft_lot) as MaxSqFootage
      FROM kc_housing
      GROUP BY condition
      ORDER BY MaxSqFootage DESC")
```

While this may work to give us our answer, technically we don't have the max. 
We have a table with 5 rows. If we want a single maximum, we can use
nested queries for this. 

Now that we have aggregated by groups, we can return the maximum of these
aggregates by nesting the previous query in another select max() statement.

```{r}
sqldf("SELECT grouped.condition, max(grouped.MaxSqFootage) as HighestMax
       FROM 
           (SELECT condition, MAX(sqft_lot) as MaxSqFootage
           FROM kc_housing
           GROUP BY condition
           ORDER BY MaxSqFootage DESC) as grouped")
```
Here you can see we actually encased our previous SQL statement in parentheses. 
Afterwards it is common practice to give our table results a name. We named it
'grouped.' Now in the outer layer of this nested query we can refer to these
records as a table named 'grouped'. So in the outer layer of the query we 
select the condition and find the max of our MaxSqFootage field. 

In practice, nested queries can allow you to do quite flexible analyses. 

#### Multiple joins
Here let's join some metadata about the cars from mtcars to the kc_cars_subset
table. Then we can left join the resulting table to the kc_housing table. 

```{r}
mtcars$cartype <- row.names(mtcars)
sqldf("SELECT kc_housing.id, kc_housing.price, kc_housing.bathrooms, temp.cartype, temp.mpg
       FROM kc_housing
       LEFT JOIN
           (SELECT kc_cars_subset.id, mtcars.cartype, mtcars.mpg
            FROM kc_cars_subset 
            INNER JOIN mtcars
            ON kc_cars_subset.cartype = mtcars.cartype) as temp
       ON kc_housing.id = temp.id
       LIMIT 30")
```


####Case Statements

We will wrap up with CASE statements, which allow you to create new variables
using if/else logic inside of SQL.

Suppose we wanted to create a new binary variable called 'BigLot' based on
whether the lot size square footage exceeds 10,000 feet. 

We can use the following statement to create the variable. 

```{r}
sqldf("SELECT id, price, sqft_lot,
       CASE WHEN sqft_lot >= 10000 THEN 1
            ELSE 0 END AS BigLot
       FROM kc_housing
       LIMIT 15")
```

Finally, we can use case statements to make new variables that we can use
in nested SQL statements, group by statements, etc. 

Suppose we wanted to take the previous query and now perform aggregation
based on the BigLot status. 

```{r}
sqldf("SELECT avg(price) as AvgPrice,
       CASE WHEN sqft_lot >= 10000 THEN 1
            ELSE 0 END AS BigLot
       FROM kc_housing
      GROUP BY BigLot")
```

As you can see case statements can provide you with some flexibile analysis. 

## Summary

In this guide, we demonstrated beginner, intermediate, and more advanced
uses of Structured Query Language (SQL). 